Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-10-02 appendix.tex
keep unambiguous states in order and add file paths to doc
preserve tip state order and add file paths
measure zero and clarifications

We show that solutions of this form never obtain $\hat{t} = t^*$ except in cases of zero or infinite branch length.
Given Table~\ref{tab:gen-sitepatprob}, all solutions to $\hat{x}_1$ have the form
\[
\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$, and the only cases to possibly obtain $\hat{x}_1=x^*$ are when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length; the same is true for $x_2$.
A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent when $(y^*)^2=y^*$, i.e., $y^*=0$ or $y^*=1$.
\end{proof}

@@

\begin{figure}
    \begin{minipage}{\linewidth}
        \centering
        \includegraphics[width=.95\textwidth]{analytic-anc-state}
        \label{ }
    \end{minipage}
    \begin{minipage}{\linewidth}
        \centering
        Maximal ancestral state split definitions
        \begin{tabular}{l}
        \hline
        $\hat{\boldsymbol\xi}=\{\xi_{\emptyset}, \xi_{1}, \xi_{2}, \xi_{3}, \xi_{123}, \xi_{12}, \xi_{23}, \xi_{13}\}$\\
        \hline
        $\hat{\boldsymbol\xi}_1=\{\emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset, \emptyset\}$\\
        $\hat{\boldsymbol\xi}_2=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \emptyset, \emptyset\}$\\
        $\hat{\boldsymbol\xi}_3=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1\}, \emptyset\}$\\
        $\hat{\boldsymbol\xi}_4=\{\emptyset, \emptyset, \emptyset, \emptyset, \{1,2\}, \emptyset, \{1,\}, \{1,2\}\}$\\
        $\hat{\boldsymbol\xi}_5=\{\emptyset, \emptyset, \emptyset, \{1\}, \{1\}, \emptyset, \{1\}, \{1\}\}$\\
        \hline
        \end{tabular}
    \end{minipage}
\caption{
Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
}
\label{fig:max-anc-state}
\end{figure}

@@

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{empirical-marginal}
\caption{
    Estimates for $\hat{p}_w$ when computing $(\hat{x}_1, \hat{y}_1, \hat{x}_2, \hat{y}_2, \hat{w})$ using L-BFGS-B optimizing the classical integrated likelihood \eqref{eq:marginal_likelihood} rather than a joint optimization procedure.
}
\label{fig:bl-general-marginal}
\end{figure}
%
%\begin{figure}
%\centering
%\includegraphics[width=.95\textwidth]{empirical-incons-intuition}
%\caption{
%Inconsistency in joint inference with intuitive curves.
%The white curve is our condition on the gradients.
%The blue curve is a condition on $x^*$ and $y^*$ such that $\emptyset$ is the most likely ancestral state.
%}
%\label{fig:empirical-incons-intuition}
%\end{figure}

>>>>

We show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
Given Table~\ref{tab:gen-sitepatprob}, all solutions $\hat{x}_1$ have the form
\[
\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$.
It is possible to obtain $\hat{x}_1=x^*$, such as when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length, but we cannot have $\hat{x}_1=x^*$ for all $x^*$.
The same is true for $x_2$.
A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
\end{proof}

Examples of generating parameters resulting in consistency of joint inference occur in the following.
Let $x^*$ and $y^*$ be such that the estimate for $\hat{x}_1$ is
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{123} + p_{13}\right) - 1 = (x^*)^2.
\]
There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
In this case, $\hat{x}_1 = x^*$ implies consistency only when $x^*=0$ or $x^*=1$.
Another case,
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{3} + p_{123} + p_{23} + p_{13}\right) - 1 = \frac{1}{2} + \frac{1}{2}(x^*)^2
\]
shows $\hat{x}_1 = x^*$ implies consistency only when $x^*=1$.

@@

\begin{figure}
    \begin{minipage}{\linewidth}
        \centering
        %file: plotting/_output/invfels/ancestral_state_conditions/non-bias/plot.png
        \includegraphics[width=.95\textwidth]{analytic-anc-state}
        \label{ }
    \end{minipage}
    \begin{minipage}{\linewidth}
        \centering
        Maximal ancestral state split definitions
        \begin{tabular}{l}
        \hline
        $\hat{\boldsymbol\xi} = \{\xi_{\emptyset},\xi_{1},\xi_{2},\xi_{3},\xi_{123},\xi_{12},\xi_{23},\xi_{13}\}$\\
        \hline
        $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$\\
        $\hat{\boldsymbol\xi}_2 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\emptyset\}$\\
        $\hat{\boldsymbol\xi}_3 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\emptyset,\{1\}\}$\\
        $\hat{\boldsymbol\xi}_4 = \{\emptyset,\emptyset,\emptyset,\emptyset,\{1,2\},\emptyset,\{1,2\},\{1\}\}$\\
        $\hat{\boldsymbol\xi}_5 = \{\emptyset,\emptyset,\emptyset,\{1\},\{1\},\emptyset,\{1\},\{1\}\}$\\
        \hline
        \end{tabular}
    \end{minipage}
\caption{
Regions of maximal ancestral state splits on the InvFels tree $\tau_1$.
}
\label{fig:max-anc-state}
\end{figure}

@@

The boundaries determining maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are obtained through maximizing over 81 separate functional values (Table~\ref{tab:likelihoods}).
Using the notation in the proof of Theorem~\ref{thm:bl_theorem}, we see that, in general, our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
\begin{align}
\ell_{\tau^*,t^*}(\tau_1, \hat{t}; \boldsymbol\xi) &= \frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1)\nonumber\\
&\qquad + \frac{1+\hat{y}_1}{2}\log(1+\hat{y}_1) + \frac{1-\hat{y}_1}{2}\log(1-\hat{y}_1)\nonumber\\
&\qquad + \frac{1+\hat{x}_2}{2}\log(1+\hat{x}_2) + \frac{1-\hat{x}_2}{2}\log(1-\hat{x}_2)\nonumber\\
&\qquad + \frac{1+\hat{y}_2}{2}\log(1+\hat{y}_2) + \frac{1-\hat{y}_2}{2}\log(1-\hat{y}_2)\nonumber\\
&\qquad + \frac{1+\hat{w}}{2}\log(1+\hat{w}) + \frac{1-\hat{w}}{2}\log(1-\hat{w}).
\end{align}
For example, in the lower right region of Fig.~\ref{fig:max-anc-state}, the maximal ancestral state split is $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ so that
\[
\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \{x^*(y^*)^2,(y^*)^2,x^*(y^*)^2,1,1\}
\]
and the exact, closed form value of the likelihood is
\begin{align}
\ell_{U,1}(x^*,y^*) &= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
&\qquad + \frac{1+(y^*)^2}{2}\log(1+(y^*)^2) + \frac{1-(y^*)^2}{2}\log(1-(y^*)^2)\nonumber\\
&\qquad + 2\log(2).
\end{align}
We obtain similar maximal values as functions of $x^*$ and $y^*$ for each ancestral state split.
The curves delineating maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are determined by these 81 functions.

@@

\begin{figure}
\centering
%file: plotting/_output/invfels/marginal_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-marginal}
\caption{
    Estimates for $\hat{p}_w$ when computing $(\hat{x}_1, \hat{y}_1, \hat{x}_2, \hat{y}_2, \hat{w})$ using L-BFGS-B optimizing the classical integrated likelihood \eqref{eq:marginal_likelihood} rather than a joint optimization procedure.
}
\label{fig:bl-general-marginal}
\end{figure}

### 2018-10-02 joint_inf.tex
preserve tip state order and add file paths
measure zero and clarifications

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
For branch length estimation on the correct topology, we show that this joint method cannot estimate consistent branch lengths except in degenerate cases, and we provide extensive empirical results for outlining the consistent bias in this setting.
\end{abstract}

@@

In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
We show that, when the topology is known and fixed, joint inference cannot be consistent except in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.

@@

When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.

\begin{restatable}{theorem}{blInconsist}
\label{thm:bl_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
For all $0 < x^*, y^* < 1$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property $\hat{t} \neq t^*$.
\end{restatable}

In words, the joint estimation procedure never recovers the true generating $t^*$ except in cases of zero or infinite branch length.
This is apparent given Table~\ref{tab:gen-sitepatprob}, as the solution $\hat{t}$ is a linear combination of $p_{\siteSplit_j}$ values, and no generating probability contains an $x^*$ or $y^*$ term.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Data generated as in Fig.~\ref{fig:max-anc-state}.
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{empirical-bias}
\caption{
Bias in branch length estimation.
Even in regions with short branch length ($p_{x^*}, p_{y^*} \le .1$) where joint optimization should perform well, there is systematic bias toward shorter branch lengths.
}
\label{fig:empirical-bias}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
In addition, we have shown there are no nontrivial generating parameters that yield consistent branch length estimates.

>>>>

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
We show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates a multifurcating topology in the limit of infinite-length sequences by estimating one or more branches to be zero length.
For branch length estimation on the correct topology, we show that this joint method cannot estimate consistent branch lengths except on a set of measure zero, and we provide extensive empirical results for outlining the consistent bias in this setting.
\end{abstract}

@@

In this paper, we show that jointly inferring trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data generated on a four-taxon tree: we compute closed form solutions to the joint objective function and demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give a multifurcating tree in the case of perfect sequence data with an infinite number of sites by estimating one or more branch lengths to be zero.
We show that, when the topology is known and fixed, joint inference cannot be consistent except on a set of measure zero, such as in cases of zero or infinite branch length, and we find similar areas through empirical means where joint inference consistently underestimates interior branch lengths.

@@

When the topology is known and fixed and we estimate only branch lengths, we show the following, i.e., that for almost all $x^*$ and $y^*$ in $(0,1)$ any branch length estimate is consistently biased.

\begin{restatable}{theorem}{blInconsist}
\label{thm:bl_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
For $0 < x^*, y^* < 1$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \max_{\boldsymbol\xi} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property $\hat{t} \neq t^*$ everywhere except a set of measure zero.
\end{restatable}

In words, the joint estimation procedure never recovers the true generating $t^*$ almost everywhere in the space of generating parameters.
This is apparent given Table~\ref{tab:gen-sitepatprob}, as the solution $\hat{t}$ is a linear combination of $p_{\siteSplit_j}$ values, and no generating probability contains an $x^*$ or $y^*$ term.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.
Thus, inconsistency is guaranteed for almost all $x^*$ and $y^*$, though not all $x^*$ and $y^*$.

@@

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Data generated as in Fig.~\ref{fig:max-anc-state}.
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/bias/plot.svg
\includegraphics[width=.95\textwidth]{empirical-bias}
\caption{
Bias in branch length estimation.
Even in regions with short branch length ($p_{x^*}, p_{y^*} \le .1$) where joint optimization should perform well, there is systematic bias toward shorter branch lengths.
}
\label{fig:empirical-bias}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, we have shown there are no sets of generating parameters other than ones of measure zero that yield consistent branch length estimates given the fixed, correct topology.
In addition, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches, which can be considered as a multifurcating topology.
Also, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
